Supplementary Materials:\\Evidence for black-swan events in animal populations

### 2016-02-05 text/ms.tex
Incorporate Nick and Trevor's comments

\textbf{Title: Evidence for black-swan events in animal populations}

@@

\textbf{Abstract}:
Black swans are improbable events that nonetheless occur---often with profound
consequences. While physical extremes, such as monsoons and heat waves, are
widely studied and increasing in magnitude and frequency, it remains unclear
the extent to which ecological populations buffer or suffer from such
extremes. Here we estimate the frequency and direction of black-swan events in
609 animal populations. We find strong evidence for black-swan events
occurring in \(\sim\)\overallBasePerc \% of populations: most frequently for
birds (\birdPH \%), mammals (\mammalsPH \%), and insects (\insectsPH \%).
Black-swan events are predominantly population crashes (\percBSDown \%),
driven by extrinsic perturbations such as climate or severe winters, and
ignoring them may underestimate extinction risk. Our results demonstrate the
importance of allowing for heavy-tailed downward events in population
modelling and developing robust conservation strategies, particularly in the
face of increased climate extremes.

\textbf{One Sentence Summary}: Black swans---statistically improbable events
with profound consequences---can and do occur in animal populations and
ignoring them may severely underestimate extinction risk.

@@

\textbf{Main text}:
Major surprises happen more often than expected in financial, social, and natural systems\cite{taleb2007, sornette2009, may2008}. Massive unpredictable market swings are responsible for the majority of financial gains and losses\cite{taleb2007}, fatalities from the largest wars dwarf those from other wars\cite{newman2005}, and the frequency of the most damaging earthquakes has exceeded past expectations\cite{sornette2009}. In ecology, background rates of global extinction are punctuated by mass extinction\cite{harnik2012}, viruses can mutate suddenly to infect new hosts, and billions of animals can die at once in mass mortality events\cite{fey2015}. Indeed, catastrophes may be the most important element affecting population persistence\cite{mangel1994} and their importance is likely to increase given projected increases in the frequency and magnitude of climate-related extremes\cite{ipcc2012}. Despite this compelling anecdotal evidence for ecological black swans, systematic evidence in population time series of higher taxa has been elusive\cite{keitt1998, allen2001, halley2002} and the vast majority of model fitting and risk forecasting assumes that uncertainty can be represented by a normal distribution\cite{brook2006a, knape2012}. If black swans occur, though, a normal distribution would under-estimate the probability of extreme events occurring.

Here, we assess the frequency and magnitude of black-swan dynamics across time series of 609 populations from a wide array of taxonomic groups---mostly birds, mammals, insects, and fishes (table~S1) \cite{SOM}. We then identify characteristics of time series or life-history traits associated with the detection of black-swan events and verify known causes. To accomplish this, we develop a framework for identifying heavy-tailed (black-swan) process noise in population dynamics, i.e.~whether the largest stochastic jumps in log abundance from one time step to the next are more extreme than typically seen with a normal distribution. Our framework allows for a range of population dynamic models, can incorporate observation uncertainty and skewness in process noise, and can be readily applied to abundance time series.

Specifically, we fit population dynamics models in which the process noise is drawn from a Student-t distribution. By estimating the degrees of freedom parameter, \(\nu\), we can estimate the degree to which the process deviations have heavy tails and are therefore evidence of black-swan events (Fig.~1, fig.~S\ref{fig:priors}). Lower values of \(\nu\) result in heavy-tailed distributions. For example, at \(\nu = 3\), the probability of drawing a value more than five standard deviations below the mean is moderately high (\(0.02\)), whereas the probability of drawing such a value from a normal distribution is tiny (\(2.9\cdot10^{-7}\)). As \(\nu\) approaches infinity, the distribution approaches the normal distribution (Fig.~1).

We found strong, but rare, evidence for black-swan dynamics: most frequently for birds (7\%) followed by mammals (5\%), and insects (3\%) (Fig.~2). Black-swan events were taxonomically widespread, occurring in 38\% of taxonomic orders. Accounting for time series length and partially pooling inference across taxonomic class and order with a hierarchical model, we found stronger evidence for black swans in insect populations than these statistics suggest---four of eight orders with the highest median probability of heavy tails were insect orders (Fig.~3a).

@@

For model fits to the population data, the probability of detecting black-swan dynamics was positively related to time-series length and negatively related to magnitude of process noise but not clearly related to population growth rate, density dependence, or maximum lifespan (Fig.~3b, fig.~S\ref{fig:correlates}). Longer time-series length was the strongest covariate of observing black-swan dynamics: there is a 1.6 times greater probability of detecting a black swan event in populations with 60 time steps than in one with 30 time steps (fig.~S\ref{fig:correlates}).

We examined all time series with published explanations of why black-swan events occurred (table~S2). The majority (\percBSDown \%) were downward events and involved a combination of multiple factors. For example, a synchronization of environmental- and predation-mediated population cycles is thought to have caused a downward black-swan event for a water vole (\emph{Arvicola terrestris}) population\cite{saucy1994}. Other black swans were the result of a sequence of extreme climate events on their own. For instance, severe winters in 1929, 1940--1942, and 1962--1963 were associated with black-swan downswings in grey heron (\emph{Ardea cinerea}) abundance in the United Kingdom\cite{stafford1971} (Fig.~4c). Our analysis finds that the last event was a combination of two heavy-tailed events in a row and the population took three times longer to recover than predicted\cite{stafford1971}. Downward black swans were sometimes followed by upward black swans. For example, during a period of population crowding and nest shortages, a population of European shag cormorants (\emph{Phalacrocorax aristotelis}) on the Farne Islands, United Kingdom declined suddenly following a red tide event in 1968\cite{potts1980}. This freed quality nest sites for first-time breeders, productivity increased, and the population experienced a rapid upswing in abundance\cite{potts1980}.

Given the prevalence of downwards events, we refit our heavy-tailed models to measure the degree of skewness of the process noise using a skew-t distribution (\(\gamma\); Fig.~4a), and used these models to make near-term risk forecasts. Aggregated across populations with strong evidence of heavy tails (median \(\nu < 10\)), 85\% of the \(\gamma\) probability density was below 1, indicating strong evidence of downwardly skewed process noise (Fig.~4b, fig.~S\ref{fig:skew-nu}). In contrast, populations that did not have heavy tails (median \(\nu \geq 70\)) had little evidence of skewed process noise (Fig.~4b; 90\% of 95\% credible intervals overlapped 1). Projecting these heavy-tailed populations forward five years revealed that assuming the standard normal process noise underestimated risk by \crashUnderRange-fold (interquartile range of 99\% lower credible interval of abundance; Fig.~4c, fig.~S\ref{fig:skew-projections}). Thus our analysis provides strong evidence for downward-skewed heavy-tailed events in abundance time series of higher taxa, and ignoring these events will tend to underestimate the risk of population declines.

There are a number of possible causes of black swan events, including unmodelled intrinsic properties of populations or extrinsic forces acting on populations. For example, we could observe black-swan dynamics if we miss an underlying mixture of processes, since a mixture of normal distributions with different variances can generate a t distribution with heavy tails\cite{allen2001}. In ecology, these processes could arise through an aggregation of populations across space, population diversity\cite{schindler2010}, or an intrinsic change in population variability through time\cite{carpenter2006}. Extrinsic ecological forces could also cause black-swan dynamics\cite{nunez2012} including extreme climate\cite{meehl2004, katz2005, ipcc2012}, predation from (or competition with) other species experiencing black swans, or sharp changes in human pressures such hunting, fishing, or habitat destruction. Alternatively, the synchrony of multiple extrinsic forces could give rise to black-swan dynamics through synergistic interactions\cite{kirby2009} or a rare alignment of non-synergistic forces\cite{denny2009}.

In light of our findings, we suggest that ecological resource management can learn from disciplines that focus on heavy tails. For example, earthquake preparedness and response is focused on black-swan events. Similarly to ecological black swans, we can rarely predict the specific timing of large earthquakes. But, earthquake preparedness involves spatial planning based on forecast probabilities to focus early detection efforts and develop disaster response plans\cite{nrc2007}. The presence of black swans also suggests that we develop management policy that is robust to heavy tails. For instance, setting target population abundances far from critical limits may buffer black-swan events\cite{caddy1996}, and maintaining genetic, phenotypic, and behavioural diversity may allow some components of populations to persist when others are affected by disease or extreme environmental forces\cite{schindler2010}. Finally, surprising, or counterintuitive ecological dynamics offer a tremendous opportunity to learn about ecological systems, evaluate when models break down, and adjust future management policy\cite{doak2008, lindenmayer2010}.

@@

\textbf{Fig.~1: Illustration of population dynamic models that allow for heavy tails.} (\textbf{A} to \textbf{B}) Probability density for the Student-t distribution with scale parameter of 1 and different values of \(\nu\). Small values of \(\nu\) create heavy tails. As \(\nu\) approaches infinity the distribution approaches the normal distribution. (\textbf{C} to \textbf{E}) Simulated population dynamics from a Gompertz model with process noise drawn from Student-t distributions with three values of \(\nu\). Coloured dots in panels \textbf{C} and \textbf{D} represent jumps with less than a 1 in 1000 chance of occurring in a normal distribution. (\textbf{F} to \textbf{H}) Estimates of \(\nu\) from models fit to the times series in \textbf{C} to \textbf{E}. Shown are posterior samples (histograms), median and interquartile range of the posterior (IQR, dots and line segments), and the exponential prior on \(\nu\) (dashed lines). Colour shading behind \textbf{F} to \textbf{H} illustrates the approximate region of heavy tails.

@@

\textbf{Fig.~4: Heavy-tailed process noise tends to be downward skewed and ignoring this can underestimate risk.} (\textbf{A}) Illustrations of Student-t distributions with 3 levels of skewness (\(\gamma\)) and heavy-tails (\(\nu\)). (\textbf{B}) Posterior density of the skewness parameters aggregated across populations grouped into heavy-tailed (\(\hat{\nu} < 10\)), slightly heavy-tailed (\(10 \leq \hat{\nu} \leq 70\)), and normal-tailed populations (\(\hat{\nu} > 70\)). Approximate mid-values from \textbf{B} are illustrated in \textbf{A}. (\textbf{C}) Example time series of heavy-tailed populations with different levels of skewness. Red dots highlight likely heavy-tailed events. Forecasts (grey regions) show median (solid lines) and lower 99\% credible intervals (dotted lines) of abundance. Black and red lines indicate forecasts from Gompertz models with lognormal and skew-t process noise, respectively.

>>>>

\textbf{Title: Black-swan events in animal populations}

@@

\textbf{Abstract}:
Black swans are improbable events that nonetheless occur---often with profound consequences. While physical extremes, such as monsoons and heat waves, are widely studied and increasing in magnitude and frequency, it remains unclear the extent to which ecological population numbers buffer or suffer from such extremes. Here we estimate the frequency and direction of black-swan events in 609 animal populations, finding strong evidence for their occurrence in \(\sim\)\overallBasePerc \% of populations. Black-swan events occur most frequently for birds (\birdPH \%), mammals (\mammalsPH \%), and insects (\insectsPH \%), and manifest primarily as   population die-offs and crashes (\percBSDown \%) driven by extrinsic perturbations from climate variability. Thus, ignoring black-swan events will under-estimate extinction risk. Our results demonstrate the importance of both modeling heavy-tailed downward events in populations, and developing conservation strategies that are robust to future increases in climate extremes.

\textbf{One Sentence Summary}: Black swans---statistically improbable events with profound consequences---occur in animal populations and understanding them will improve climate risk assessments.

@@

\textbf{Main text}:
Major surprises happen more often than expected in financial, social, and natural systems\cite{taleb2007, sornette2009, may2008}. Massive unpredictable market swings are responsible for the majority of financial gains and losses\cite{taleb2007}, fatalities from the largest wars dwarf those from all others \cite{newman2005}, and the frequency of the most damaging earthquakes has exceeded past expectations\cite{sornette2009}. In ecological systems, background rates of global extinction are punctuated by mass extinction\cite{harnik2012}, viruses can mutate suddenly to infect new hosts, and billions of animals can die at once in mass mortality events\cite{fey2015}. Indeed, such die-offs may be the most important element affecting population persistence\cite{mangel1994} and their importance is likely to increase given projected increases in the frequency and magnitude of climate-related extremes\cite{ipcc2012}. Despite this anecdotal evidence for ecological black swans, systemtic searches of large numbers of timeseries have yet to reveal black-swan events\cite{keitt1998, allen2001, halley2002} However, the overwhelming majority of model fitting and risk forecasting assumes that deviations from model predictions can be represented by a normal distribution\cite[e.g.][]{brook2006a, knape2012}. If black swans occur, though, a normal distribution would under-estimate the probability of extreme events occurring.

Here, we develop a new approach to estimate the frequency and magnitude of black-swan dynamics across time series of 609 populations from a wide array of taxonomic groups---including many  birds, mammals, insects, and fishes (table~S1) \cite{SOM}. We identify characteristics of time series or life-history traits associated with the detection of black-swan events and verify known causes. To accomplish this, we develop a framework for identifying heavy-tailed (black-swan) process noise in population dynamics, i.e.~whether the largest stochastic jumps in log abundance from one time step to the next are more extreme than typically seen with a normal distribution. Our framework allows for a range of population dynamic models, can incorporate observation uncertainty and skewness in process noise, and can be readily applied to abundance time series.

Specifically, we fit population dynamics models in which the process noise is drawn from a Student-t distribution. By estimating the degrees of freedom parameter, \(\nu\), we can estimate the degree to which the process deviations have heavy tails and are therefore diagnositc of black-swan events (Fig.~1, fig.~S\ref{fig:priors}). Lower values of \(\nu\) result in heavy-tailed distributions. For example, at \(\nu = 2\), the probability of drawing a value more than five standard deviations below the mean is almost 70,000 times higher than the probability of drawing such a value from a normal distribution (\(0.02\) vs.\ \(2.9\cdot10^{-7}\)). As \(\nu\) approaches infinity, the distribution approaches the normal distribution (Fig.~1).

We found evidence that black-swan dynamics are infreqent but highly influentiual: they occur most frequently for birds (7\%), mammals (5\%), and insects (3\%) and almost never in fishes for this dataset (Fig.~2). Black-swan events were taxonomically widespread, occurring in 38\% of taxonomic orders. Accounting for time series length and partially pooling inference across taxonomic class and order with a hierarchical model, we found stronger evidence for black swans in insect populations than these statistics suggest---four of eight orders with the highest median probability of heavy tails were insect orders (Fig.~3a).

@@

For model fits to the population data, the probability of detecting black-swan dynamics was positively related to time-series length and negatively related to magnitude of process noise but not clearly related to population growth rate, density dependence, or maximum lifespan (Fig.~3b, fig.~S\ref{fig:correlates}). Longer time-series length was the strongest covariate of observing black-swan dynamics: there is a 1.6 times greater probability of detecting a black-swan event in populations with 60 time steps than in one with 30 time steps (fig.~S\ref{fig:correlates}).

The majority of black-swan events (\percBSDown \%) were downward (die-offs) rather than upward (unexpectedly rapid abundance increases). Of the black-swan events with published explanations (table~S2), the majority involved a combination of multiple factors. For example, a synchronization of environmental- and predation-mediated population cycles is thought to have caused a downward black-swan event for a water vole (\emph{Arvicola terrestris}) population\cite{saucy1994}. Other black swans were the result of a sequence of extreme climate events on their own. For instance, severe winters in 1929, 1940--1942, and 1962--1963 were associated with black-swan downswings in grey heron (\emph{Ardea cinerea}) abundance in the United Kingdom\cite{stafford1971} (Fig.~4c). Our analysis finds that the last event was a combination of two heavy-tailed events in a row and the population took three times longer to recover than predicted\cite{stafford1971}. Downward black swans were sometimes followed by upward black swans. For example, during a period of population crowding and nest shortages, a population of European shag cormorants (\emph{Phalacrocorax aristotelis}) on the Farne Islands, United Kingdom declined suddenly following a red tide event in 1968\cite{potts1980}. This freed up quality nest sites for first-time breeders, productivity increased, and the population experienced a rapid upswing in abundance\cite{potts1980}.

Given the prevalence of downwards events, we refit our heavy-tailed models to measure the degree of skewness \(\gamma\) of the process noise using a skew-t distribution (Fig.~4a), and used these models to make near-term risk forecasts. Aggregated across populations with strong evidence of heavy tails (median \(\nu < 10\)), 85\% of the \(\gamma\) probability density was below 1, indicating strong evidence of downwardly skewed process noise (Fig.~4b, fig.~S\ref{fig:skew-nu}). In contrast, populations that did not have heavy tails (median \(\nu \geq 70\)) had little evidence of skewed process noise (Fig.~4b; 90\% of 95\% credible intervals overlapped 1). Projecting these heavy-tailed populations forward five years revealed that assuming the standard normal process noise underestimated risk by \crashUnderRange-fold (interquartile range of 99\% lower credible interval of abundance; Fig.~4c, fig.~S\ref{fig:skew-projections}). Thus our analysis provides strong evidence for downward-skewed heavy-tailed events in abundance time series of higher taxa, and ignoring these events will tend to underestimate the risk of population declines.

% TODO How are you defining risk. See if you can make this more concrete to avoid IUCN people getting upset with lack of specificity (i.e. Resit akcakaya)

There are a number of possible causes of black-swan events, including unmodelled intrinsic properties of populations or extrinsic forces acting on populations. For example, we could observe black-swan dynamics if we miss an underlying mixture of processes, since a mixture of normal distributions with different variances can generate a t distribution with heavy tails\cite{allen2001}. In ecology, these processes could arise through an aggregation of populations across space, population diversity\cite{schindler2010}, or an intrinsic change in population variability through time\cite{carpenter2006}. Extrinsic ecological forces could also cause black-swan dynamics\cite{nunez2012} including extreme climate events\cite{meehl2004, katz2005, ipcc2012}, predation from (or competition with) other species experiencing black swans, or sharp changes in human pressures such hunting, fishing, or habitat destruction. Alternatively, the synchrony of multiple extrinsic forces could give rise to black-swan dynamics through synergistic interactions\cite{kirby2009} or a rare alignment of non-synergistic forces\cite{denny2009}.

In light of our findings, we suggest that ecological resource management can learn from disciplines that focus on heavy tails. For example, earthquake preparedness and response is focused on black-swan events. Similarly to ecological black swans, we can rarely predict the specific timing of large earthquakes. But, earthquake preparedness involves spatial planning based on forecast probabilities to focus early detection efforts and develop disaster response plans\cite{nrc2007}. The presence of black swans also suggests that we develop management policy that is robust to heavy tails. For instance, setting target population abundances far from critical limits will buffer against black-swan events\cite{caddy1996}, and maintaining genetic, phenotypic, and behavioural diversity may allow some components of populations to persist when others are affected by disease or extreme environmental forces\cite{schindler2010}. Finally, surprising, or counterintuitive ecological dynamics offer a tremendous opportunity to learn about ecological systems, evaluate when models break down, and adjust future management policy\cite{doak2008, lindenmayer2010}.

@@

\textbf{Fig.~1: Illustration of population dynamic models that allow for heavy tails.} (\textbf{A} to \textbf{B}) Probability density for the Student-t distribution with scale parameter of 1 and different values of \(\nu\). Small values of \(\nu\) create heavy tails while as \(\nu\) approaches infinity the distribution approaches the normal distribution. (\textbf{C} to \textbf{E}) Simulated population dynamics from a Gompertz model with process noise drawn from Student-t distributions with three values of \(\nu\). Coloured dots in panels \textbf{C} and \textbf{D} represent jumps with less than a 1 in 1000 chance of occurring in a normal distribution. (\textbf{F} to \textbf{H}) Estimates of \(\nu\) from models fit to the times series in \textbf{C} to \textbf{E}. Shown are posterior samples (histograms), median and interquartile range of the posterior (IQR, dots and line segments), and the exponential prior on \(\nu\) (dashed lines). Colour shading behind \textbf{F} to \textbf{H} illustrates the approximate region of heavy tails.

@@

% TODO The A, B, C are too close to the text in the figure.
\textbf{Fig.~4: Heavy-tailed process noise tends to be downward skewed and ignoring this can underestimate risk.} (\textbf{A}) Illustrations of Student-t distributions with 3 levels of skewness (\(\gamma\)) and heavy-tails (\(\nu\)). (\textbf{B}) Posterior density of the skewness parameters aggregated across populations grouped into heavy-tailed (\(\hat{\nu} < 10\)), slightly heavy-tailed (\(10 \leq \hat{\nu} \leq 70\)), and normal-tailed populations (\(\hat{\nu} > 70\)). Approximate mid-values from \textbf{B} are illustrated in \textbf{A}. (\textbf{C}) Example time series of heavy-tailed populations with different levels of skewness. Red dots highlight likely heavy-tailed events. Forecasts (grey regions) show median (solid lines) and lower 99\% credible intervals (dotted lines) of abundance. Black and red lines indicate forecasts from Gompertz models with lognormal and skew-t process noise, respectively.
